High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2019-09-05 paper/main.tex
copied newest content
finished first pass of edits

One challenge to fairly evaluating high-order correlations is that if
the kernel used in Equation~\ref{eqn:timecorr} is wider than a single
timepoint, each repeated application of the equation will result in
further temporal blur.  Because our primary assessment metric is
temporal decoding accuracy, this unfairly biases against detecting
meaningful signal in higher-order correlations (relative to
lower-order correlations).  We attempted to mitigate temporal blur in
estimating each $\mathbf{X}_n$ by using a Dirac $\delta$ function
kernel (which places all of its mass over a single timepoint;
Fig.~\ref{fig:kernels}b) to compute each lower-order correlation
($\mathbf{X}_1, \mathbf{X}_2, ..., \mathbf{X}_{n-1}$).  We then used a
new (potentially wider, as described below) kernel to compute
$\mathbf{X}_{n}$ from $\mathbf{X}_{n-1}$.  In this way, temporal
blurring was applied only in the last step of computing
$\mathbf{X}_n$.  We note that, because each $\mathbf{X}_n$ is a
low-dimensional representation of the corresponding $\mathbf{Y}_n$,
the higher-order correlations we estimated reflect true correlations
in the data with lower-fidelity than estimates of lower-order
correlations.  Thefore, even after correcting for temporal blurring,
our approach is still biased against finding meaningful signal in
higher-order correlations.

@@

Both sets of temporal decoding analyses yielded qualitatively similar
results for the auditory (non-rest) conditions of the experiment
(Fig.~\ref{fig:decoding}: pink, yellow, and teal lines).  The highest
decoding accuracy for participants who listened to the intact
(unscrambled) story was achieved using high-order dynamic correlations
(PCA: second-order; eigenvector-centrality: fourth-order).  Scrambled
versions of the story were best decoded by lower-order correlations
(PCA/paragraph: first-order; PCA/word: order zero; eigenvector
centrality/paragraph: order zero; eigenvector centrality/word: order
zero).  The two sets of analyses yielded different decoding results on
resting state data (Fig.~\ref{fig:decoding}: purple lines).  We note
that the resting state times could be decoded reliably, but only very
slightly above chance; we speculate that the decoders might have
picked up on attentional drift, boredom, or tiredness; we hypothesize
these all increase throughout the resting state scan.  Our decoders
might be picking up on aspects of these loosely defined cognitive
states that are common across individuals.  The PCA-based approach
achieved the highest resting state decoding accuracy using order zero
features (non-correlational activation-based), whereas the eigenvector
centrality-based approach achieved the highest resting state decoding
accuracy using second-order correlations.  Taken together, these
analyses indicate that high-level cognitive processing (while
listening to the intact story) is reflected in the dynamics of
high-order correlations in brain activity, whereas lower-level
cognitive processing (while listening to scrambled versions of the
story that lack rich meaning) is reflected in the dynamics of
lower-order correlations and non-correlational activity dynamics.
Further, these patterns are associated both with the underlying
activity patterns (characterized using PCA) and also with the changing
relative positions that different brain areas occupy in their
associated networks (characterized using eigenvector centrality).

@@

For all of the story listening conditions (intact, paragraph, and
word), we found that first- and second-order correlations were most
strongly associated with auditory and speech processing.  During
intact story listening, third-order correlations reflected integration
with visual areas, and fourth-order correlations reflected integration
with areas associated with high-level cognition and cognitive control,
such as the ventrolateral prefrontal cortex.  However, during
listening to temporally scrambled stories, these higher-order
correlations instead involved interactions with additional regions
associated with speech and semantic processing.  By contrast, we found
a much different set of patterns in the resting state data.
First-order resting state correlations were most strongly associated
with regions involved in counting and numerical understanding.
Second-order resting state correlations were strongest in visual
areas; third-order correlations were strongest in task-positive areas;
and fourth-order correlations were strongest in regions associated
with autobiographical and episodic memory.  We carried out analogous
analyses to create maps (and decode the top associated Neurosynth
terms) for up to fifteenth-order correlations (Figs. S1, S2, S3, and
S4).  Of note, examining fifteenth-order correlations between 700
nodes using conventional methods would have required storing roughly
$\frac{700^{2 \times 15}}{2} \approx 1.13 \times 10^{85}$ floating point
numbers-- assuming single-precision (32 bits each), this would require
roughly 32 times as many bits as there are molecules in the known
universe!  Although these fifteenth-order correlations do appear
(visually) to have some well-formed structure, we provide this latter
example primarily as a demonstration of the efficiency and scalability
of our approach.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/discussion}
  \caption{\textbf{Proposed high-order network dynamics underlying
      high-level cognition.}
    Higher orders of network interactions support higher-level aspects
    of cognitive processing.  When tasks evoke richer, deeper, and/or
    higher-level processing, this is reflected in higher-order network
    interactions.}
  \label{fig:discussion}
\end{figure}

The notion that cognition is reflected in (and possibly mediated by)
patterns of network dynamics has been suggested by or proposed in
myriad empirical studies and reviews~\citep[e.g.,][]{DemeEtal19,
  Turk13, LuriEtal18, FongEtal19, ParkEtal18b, PretEtal17,
  RoyEtal19, LiegEtal19, ZouEtal19, ChanGlov10, GonzEtal19}.  Our
study extends this line of work by finding cognitively relevant
higher-order network dynamics that reflect ongoing cognition.  Our
findings complement other work that uses graph theory and topology to
characterize how brain networks reconfigure during
cognition~\citep[e.g.,][]{BassEtal06, ZhenEtal19, McInJirs19,
  TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.

@@

Another potential limitation of our approach relates to recent work
suggesting that the brain undergoes rapid state changes, for example
across event boundaries~\citep[e.g.,][]{BaldEtal17}.
\cite{ShapEtal19} used hidden semi-Markov models to estimate
state-specific network dynamics.  Our general approach might be
extended by considering putative state transitions. For example,
rather than weighting all timepoints using a similar kernel
(Eqn.~\ref{eqn:timecorr}), the kernel function could adapt on a
timepoint-by-timepoint basis such that only timepoints determined to
be in the same ``state'' were given non-zero weight.

@@

\subsection*{Concluding remarks}
The complex hierarchy of dynamic interactions that underlie our
thoughts is perhaps the greatest mystery in modern science.  Methods
for characterizing the dynamics of high-order correlations in neural
data provides a window into the neural basis of cognition.
\textbf{Need a good closing sentence.}

>>>>

One challenge to fairly evaluating high-order correlations is that if
the kernel used in Equation~\ref{eqn:timecorr} is wider than a single
timepoint, each repeated application of the equation will result in
further temporal blur.  Because our primary assessment metric is
temporal decoding accuracy, this unfairly biases against detecting
meaningful signal in higher-order correlations (relative to
lower-order correlations).  We attempted to mitigate temporal blur in
estimating each $\mathbf{X}_n$ by using a Dirac $\delta$ function
kernel (which places all of its mass over a single timepoint;
Fig.~\ref{fig:kernels}b) to compute each lower-order correlation
($\mathbf{X}_1, \mathbf{X}_2, ..., \mathbf{X}_{n-1}$).  We then used a
new (potentially wider, as described below) kernel to compute
$\mathbf{X}_{n}$ from $\mathbf{X}_{n-1}$.  In this way, temporal
blurring was applied only in the last step of computing
$\mathbf{X}_n$.  We note that, because each $\mathbf{X}_n$ is a
low-dimensional representation of the corresponding $\mathbf{Y}_n$,
the higher-order correlations we estimated reflect true correlations
in the data with lower-fidelity than estimates of lower-order
correlations.  Therefore, even after correcting for temporal blurring,
our approach is still biased against finding meaningful signal in
higher-order correlations.

@@

Both sets of temporal decoding analyses yielded qualitatively similar
results for the auditory (non-rest) conditions of the experiment
(Fig.~\ref{fig:decoding}: pink, yellow, and teal lines).  The highest
decoding accuracy for participants who listened to the intact
(unscrambled) story was achieved using high-order dynamic correlations
(PCA: second-order; eigenvector-centrality: fourth-order).  Scrambled
versions of the story were best decoded by lower-order correlations
(PCA/paragraph: first-order; PCA/word: order zero; eigenvector
centrality/paragraph: order zero; eigenvector centrality/word: order
zero).  The two sets of analyses yielded different decoding results on
resting state data (Fig.~\ref{fig:decoding}: purple lines).  We note
that while the resting state times could be decoded reliably, the
accuracies were only very slightly above chance.  We speculate that
the decoders might have picked up on attentional drift, boredom, or
tiredness; we hypothesize these all increase throughout the resting
state scan.  The decoders might be picking up on aspects of these
loosely defined cognitive states that are common across individuals.
The PCA-based approach achieved the highest resting state decoding
accuracy using order zero features (non-correlational,
activation-based), whereas the eigenvector centrality-based approach
achieved the highest resting state decoding accuracy using
second-order correlations.  Taken together, these analyses indicate
that high-level cognitive processing (while listening to the intact
story) is reflected in the dynamics of high-order correlations in
brain activity, whereas lower-level cognitive processing (while
listening to scrambled versions of the story that lack rich meaning)
is reflected in the dynamics of lower-order correlations and
non-correlational activity dynamics.  Further, these patterns are
associated both with the underlying activity patterns (characterized
using PCA) and also with the changing relative positions that
different brain areas occupy in their associated networks
(characterized using eigenvector centrality).

@@

For all of the story listening conditions (intact, paragraph, and
word), we found that first- and second-order correlations were most
strongly associated with auditory and speech processing areas.  During
intact story listening, third-order correlations reflected integration
with visual areas, and fourth-order correlations reflected integration
with areas associated with high-level cognition and cognitive control,
such as the ventrolateral prefrontal cortex.  However, during
listening to temporally scrambled stories, these higher-order
correlations instead involved interactions with additional regions
associated with speech and semantic processing.  By contrast, we found
a much different set of patterns in the resting state data.
First-order resting state correlations were most strongly associated
with regions involved in counting and numerical understanding.
Second-order resting state correlations were strongest in visual
areas; third-order correlations were strongest in task-positive areas;
and fourth-order correlations were strongest in regions associated
with autobiographical and episodic memory.  We carried out analogous
analyses to create maps (and decode the top associated Neurosynth
terms) for up to fifteenth-order correlations (Figs. S1, S2, S3, and
S4).  Of note, examining fifteenth-order correlations between 700
nodes using conventional methods would have required storing roughly
$\frac{700^{2 \times 15}}{2} \approx 1.13 \times 10^{85}$ floating
point numbers-- assuming single-precision (32 bits each), this would
require roughly 32 times as many bits as there are molecules in the
known universe!  Although these fifteenth-order correlations do appear
(visually) to have some well-formed structure, we provide this latter
example primarily as a demonstration of the efficiency and scalability
of our approach.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/discussion}
  \caption{\textbf{Proposed high-order network dynamics underlying
      high-level cognition during story listening.}  Higher orders of
    network interactions support higher-level aspects of cognitive
    processing.  When tasks evoke richer, deeper, and/or higher-level
    processing, this is reflected in higher-order network
    interactions.}
  \label{fig:discussion}
\end{figure}

The notion that cognition is reflected in (and possibly mediated by)
patterns of network dynamics has been suggested by or proposed in
myriad empirical studies and reviews~\citep[e.g.,][]{DemeEtal19,
  Turk13, LuriEtal18, FongEtal19, ParkEtal18b, PretEtal17,
  RoyEtal19, LiegEtal19, ZouEtal19, ChanGlov10, GonzEtal19}.  Our
study extends this line of work by finding cognitively relevant
\textit{higher-order} network dynamics that reflect ongoing cognition.  Our
findings complement other work that uses graph theory and topology to
characterize how brain networks reconfigure during
cognition~\citep[e.g.,][]{BassEtal06, ZhenEtal19, McInJirs19,
  TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.

@@

Another potential limitation of our approach relates to recent work
suggesting that the brain undergoes rapid state changes, for example
across event boundaries~\citep[e.g.,][]{BaldEtal17}.
\cite{ShapEtal19} used hidden semi-Markov models to estimate
state-specific network dynamics~\citep[also see][]{VidaEtal18}.  Our
general approach might be extended by considering putative state
transitions. For example, rather than weighting all timepoints using a
similar kernel (Eqn.~\ref{eqn:timecorr}), the kernel function could
adapt on a timepoint-by-timepoint basis such that only timepoints
determined to be in the same ``state'' were given non-zero weight.

@@

\subsection*{Concluding remarks}
The complex hierarchy of dynamic interactions that underlie our
thoughts is perhaps the greatest mystery in modern science.  Methods
for characterizing the dynamics of high-order correlations in neural
data provides a window into the neural basis of cognition.  By showing
that high-level cognition is reflected in high-order network dynamics
we have elucidated the next step on the path towards understanding the
neural basis of cognition.
